Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-12-04 pkg/circtree/inst/wind_paper/draft/draft.Rnw
further revision based on Reto's feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2910 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: f7af6bd8b52fdf72ff99966ca9556124b490479f

% -------------------------------------------------------------------
% Probabilistic circular modeling
% -------------------------------------------------------------------
%\newpage
\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper
circular probability distribution, i.e., a probability distribution which
accounts for the periodicity of circular data. Generally, this feature can be
obtained by ``wrapping'' the probability density function of any continuous
distribution around the unit circle \citep{Mardia+Jupp:2009}. In that way, the
wrapped Cauchy distribution or the wrapped normal distribution can be employed
to model symmetric unimodal circular data. A close approximation to the wrapped
normal distribution that is mathematically simpler and hence easier to use
\citep{Fisher:1993} is provided by the von Mises distribution, a purely
circular distribution which is also known as ``the circular normal
distribution'' and is a common choice for probabilistic modeling of circular
data. Based on a location parameter $\mu \in [0, 2\pi)$ and a concentration
parameter $\kappa > 0$ the density of the von Mises distribution for an
observation $y \in [0, 2\pi)$ is given by:
%%
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
%%
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
$0$ \citep[see, e.g.,][for a more detailed
overview]{Jammalamadaka+Sengupta:2001}.

@@

%%
\begin{table}[t!]
  \caption[Table caption text]{Overview of the data sets employed in the case
    study: For Innsbruck and Vienna, various meteorological variables and derived
    quantities of these are considered at the respective stations, located either
    directly at the airports or in their vicinities.}
  \label{tab:data}
  \begin{center}
    \begin{tabular}{l  l}
      \hline
      Data components               & Description \\
      \hline
      Temporal information:         & Daytime, day of the year \\
      \hline
      Meteorological variables:     & Wind direction, wind (gust) speed, \\
                                    & (reduced) air pressure, relative humidity, \\
                                    & temperature\\ 
      \noalign{\vskip 1mm}                           
      Derived quantities:           & 3-hourly means/minima/maxima, \\
                                    & 1-hourly and 3-hourly temporal changes, \\
                                    & spatial differences towards the airport\\
      \noalign{\vskip 1mm}                              
      Weather stations (Innsbruck): & 4 stations at the airport, \\
                                    & Igls, Kematen, Kufstein, Landeck, Patscherkofel, \\
                                    & Steinach\\
      \noalign{\vskip 1mm}
      Weather stations (Vienna):    & 9 stations at the airport, \\
                                    & Arsenal, Donaufeld, Exelberg, G\"anserndorf, \\
                                    & Gro{\ss}-Enzersdorf, Gumpoldskirchen, Hohe Warte,\\
                                    & Innere Stadt, Jubil\"aumswarte, Mariabrunn, \\
                                    & Seibersdorf, Unterlaa, Wolkersdorf\\
      \hline
    \end{tabular}
  \end{center}
\end{table}
%%

@@

    For simplicity, rather to fit one single climatological model, a single model is estimated 

@@

    using the $31$~days centered around the day of interest for all available years in the training data.

@@

    Persistence describes the previous value of a single weather quantity in a time

@@

    any major change takes place in the response variable during the forecast

@@

    of the forecasting step whether persistence can provide overall reliable
    predictions.

@@

    response values according to the description in Section~\ref{sec:prob_circ}. We
    fit one model for every time point in the validation data set employing the

@@

    stronger influence of observations closer to the time of interest exponential

@@

    prediction the current observation and all previous five observations together
    have an equal influence rate of 50 percent. Observations with longer time lags
    have exponential weights below $0.01$ and are therefore omitted from the

@@

    we use the components of the lagged 2-dimensional wind vector $vec$ and the

@@

    employing the CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} with a

@@

    Section~\ref{sec:circforest}, a circular forest is constructed based on 100
    individual trees employing the R-package \pkg{circtree}. Each of these trees is
    again built by the CTree algorithm on a subsample containing a fraction of 30
    percent of the original learning data. As it can be expected that the lagged

@@

To compare the predictive performances of all proposed models a circular
analogue of the continuous ranked probability score (CRPS) as introduced by
\cite{Grimit+Gneiting+Berrocal:2006} is computed. Just as the linear version of
the CRPS \citep[for more details see][]{Hersbach:2000} it is a proper scoring
rule \citep{Gneiting+Raftery:2007} and measures the difference between an
observation and the corresponding predicted distribution function in order to
assess the probabilistic goodness of fit for the estimated model. Hence, the
lower the CRPS value the better the predictive performance. Contrary to the
linear version, the circular CRPS reduces not to the absolute error but to the
angular distance when the forecast is deterministic.

In addition to the raw CRPS, also CRPS-based skill scores are computed to
assess differences in the improvement of the various statistical models over
the climatological model used as a reference:
%%
\begin{equation} 
  \text{CRPSS}_{\text{model}} = 1 -
  \frac{\text{CRPS}_{\text{model}}}{\text{CRPS}_{\text{climatology}}}.
\end{equation}
%%

All scores presented in the next section are computed out-of-sample based on
five years of observations. For the persistency model only time points prior to
the time of interest are used and the validation is performed rolling over all
observations. For all other models a five-fold cross-validation is employed
using up to four calendar years for model training and the remaining single
calendar year for validation. Due to the large sample size of 24 hourly values
per day over five years, some kind of temporal aggregation is needed to ensure
a correct visual comparison of the individual methods. Performed analyses have
shown that for the employed models the variability of the predictive
performance over the five years is lower than over a single day or over a
single year. Hence, CRPS and CRPS skill scores are aggregated over the
respective five validation years which yields 24 hourly scores per month
averaged over the five validation years.

\subsection{Results} 
This section provides a detailed analysis on the predictive performance of the
different proposed statistical models applied to probabilistic wind direction
forecasting. To ensure a comprehensive comparison of the models, wind direction
forecasts are evaluated for two different lead times at two airports with
different climatological site characteristics.
Figure~\ref{fig:boxplot_crpsraw} shows the CRPS values of the employed models
at forecast steps 1\,h and 3\,h for the airports Innsbruck (Panels~a,\,c) and
Vienna (Panels~b,\,d). The scores are aggregated over the five validation years,
yielding yearly mean values for every hour per calendar month, with a lower
score indicating better performance. For both stations at both forecast steps,
the circular forest overall provides the best predictive performance, followed
by the circular regression tree and, except for the $3$\,h forecast at
Innsbruck, the persistency model. In comparison to the circular regression tree and
forest, for both stations and forecast steps, the climatological model and the
linear model show clearly higher CRPS values and hence a lower predictive
performance.
%%
\begin{figure}[t]
  \centering
  \setkeys{Gin}{width=1.0\textwidth}
  \includegraphics{_plot_circforest_validation_crpsraw_agg_comparison_with_lowff_v14b.pdf}
  \caption{CRPS skill scores for wind direction forecasts based on the full
    predictive von~Mises distribution for $+1$ and $+3$\,h forecasts at the
    airports Innsbruck and Vienna. Each box-and-whisker contains $24$~hourly scores
    for each of the $12$~months averaged over the $5$~validation years which
    yields a total of $288$~yearly mean values. The scores are shown for the
    climatological, persistency and the linear model as well as for the circular
    distributional tree and forest.}
  \label{fig:boxplot_crpsraw} 
\end{figure}
%%
The different site characteristics of the airports Innsbruck
(Figure~\ref{fig:boxplot_crpsraw}\,a,\,c) and Vienna
(Figure~\ref{fig:boxplot_crpsraw}\,b,\,d) seem to have an effect on the
absolute level of the model performances and on their respective predictive
performance variances. At Innsbruck, due to the surrounding mountains a
limited number of possible wind directions exists, namely the three wind
regimes discussed for Figure~\ref{fig:tree_ibk} in
Section~\ref{sec:introduction}. Therefore, for Innsbruck the wind direction
remains rather constant in one of these possible states, but once a change
takes place it is mostly a major wind direction shift. Due to the few wind
regimes the rather inflexible climatological and linear models score relatively
well with similar CRPS values as the other models
(Figure~\ref{fig:boxplot_crpsraw}\,a,\,c). In addition, at Innsbruck the
potential high prediction errors in case of a paradigm shift seem to lead to a
higher variation in the predictive performance for all models in comparison to
Vienna; this variation is especially high for the persistency model due to its
strong vulnerability to abrupt wind shifts. On the contrary, at Vienna the
geographic surroundings do not constrain any wind direction which leads to
smaller and less abrupt changes in the wind direction as well as less
pronounced wind regimes. This seems to weaken the predictive performance of the
climatological and linear models, and to reduce the performance variability for
all models (Figure~\ref{fig:boxplot_crpsraw}\,b,\,d).

The different forecast steps have apparently only a minor effect on the
predictive performance of the climatological model and the linear model at both
stations. However, for the persistency model, at both stations, higher scores
for the $3$\,h forecast (Figure~\ref{fig:boxplot_crpsraw}\,c,\,d) reveal a
lower performance for longer lead times; this is probably due to the lower 
information content of minimum 3-hourly instead of 1-hourly lagged response 
values employed as covariates in the persistency model. 
The circular regression tree and forest seem to
partially compensate for the lower skill of the lagged response values by other
covariates, hence their predictive performance only slightly decreases for the
longer lead time. This compensation is especially evident for Innsbruck, where
the performance difference between the persistency model and the tree-based
methods significantly increases from the 1-hourly to the 3-hourly forecast.

In addition to the raw CRPS (Figure~\ref{fig:boxplot_crpsraw}), CRPS skill
scores with the climatological model as a reference are provided in
Figure~\ref{fig:boxplot_crpsskill}. Skill scores are in percent, and positive
values indicate an improvement in the predictive performance over the
reference. For all setups, the circular forest has the highest skill scores
with a mean performance gain of 13--25\% and 58--71\% for Innsbruck and Vienna,
respectively. As discussed for Figure~\ref{fig:boxplot_crpsraw}, this
improvement over the climatological model is lower for Innsbruck due to the low
number of predominant wind regimes and hence a relatively good performance of
the climatological model. Additionally, Figure~\ref{fig:boxplot_crpsskill}
shows that while the persistency model's performance is lower than the
reference (Panel~c) the tree-based models can compensate for the low skill of the 
lagged response values employed as covariates and, hence, are still significantly 
superior to the reference.

@@

% -------------------------------------------------------------------
% Summary
% -------------------------------------------------------------------
\section{Summary and conclusion}
\label{sec:summary}
Circular regression trees for circular responses are established by
coupling model-based recursive partitioning with the von Mises distribution.
The resulting trees can capture non-linear changes, shifts, and potential
interactions in covariates without prespecification of such effects. This is
particularly useful for modeling a highly fluctuating response such as wind
direction. \dots

>>>>

% -------------------------------------------------------------------
% Probabilistic circular modeling
% -------------------------------------------------------------------
%\newpage
\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper
circular probability distribution, i.e., a probability distribution which
accounts for the periodicity of circular data. Generally, this feature can be
obtained by ``wrapping'' the probability density function of any continuous
distribution around the unit circle \citep{Mardia+Jupp:2009}. In that way, the
wrapped Cauchy distribution or the wrapped normal distribution can be employed
to model symmetric unimodal circular data. A close approximation to the wrapped
normal distribution that is mathematically simpler and hence easier to use
\citep{Fisher:1993} is provided by the von Mises distribution, a purely
circular distribution which is also known as ``the circular normal
distribution'' and is a common choice for probabilistic modeling of circular
data. Based on a location parameter $\mu \in [0, 2\pi)$ and a concentration
parameter $\kappa > 0$ the density of the von Mises distribution for an
observation $y \in [0, 2\pi)$ is given by:
%%
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
%%
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
$0$ (see, e.g., \citealt{Jammalamadaka+Sengupta:2001} or \citealt{Ley+Verdebout:2017} for a more detailed overview).

@@

%%
\begin{table}[t!]
  \caption[Table caption text]{Overview of the data sets employed in the case
    study: For Innsbruck and Vienna, various meteorological variables and derived
    quantities of these are considered at the respective stations, located either
    directly at the airports or in their vicinities.}
  \label{tab:data}
  \begin{center}
    \begin{tabular}{l  l}
      \hline
      Data components               & Description \\
      \hline
      Temporal information:         & Daytime, day of the year \\
      \noalign{\vskip 1mm}   
      Meteorological variables:     & Wind direction, wind (gust) speed, \\
                                    & (reduced) air pressure, relative humidity, \\
                                    & temperature\\ 
      \noalign{\vskip 1mm}                           
      Derived quantities:           & 3-hourly means/minima/maxima, \\
                                    & 1-hourly and 3-hourly temporal changes, \\
                                    & spatial differences towards the airport\\
      \noalign{\vskip 1mm}                              
      Weather stations (Innsbruck): & 4 stations at the airport, \\
                                    & Igls, Kematen, Kufstein, Landeck, Patscherkofel, \\
                                    & Steinach\\
      \noalign{\vskip 1mm}
      Weather stations (Vienna):    & 9 stations at the airport, \\
                                    & Arsenal, Donaufeld, Exelberg, G\"anserndorf, \\
                                    & Gro{\ss}-Enzersdorf, Gumpoldskirchen, Hohe Warte,\\
                                    & Innere Stadt, Jubil\"aumswarte, Mariabrunn, \\
                                    & Seibersdorf, Unterlaa, Wolkersdorf\\
      \hline
    \end{tabular}
  \end{center}
\end{table}
%%

@@

    For simplicity, rather to fit one single climatological model, separate models are estimated 

@@

    using the $31$~days centered around the day of interest over all available years in the training data.

@@

    The persistence describes the previous value of a single weather quantity in a time

@@

    the regime changes takes place in the response variable during the forecast

@@

    of the forecasting step whether the persistence can provide overall reliable
    information.

@@

    response values according to the description in Section~\ref{sec:prob_circ}. 
    We fit one model for every hour throughout 
    the validation data set employing the

@@

    stronger influence of observations closer to the time of interest, exponential

@@

    prediction an equal influence rate of 50 percent is assigned both to the current 
    observation and to the previous five observations together. Observations with longer 
    time lags have exponential weights below $0.01$ and are therefore omitted from the

@@

    we use the components of the lagged 2-dimensional wind vector $\bm{vec} = (vec_1, vec_2)^{\top}$ and the

@@

    employing the CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} using a

@@

    Section~\ref{sec:circforest}, the circular forest used in this study is constructed 
    based on $100$~individual trees employing the R-package \pkg{circtree}. Each of these 
    trees is again built by the CTree algorithm on a subsample containing a fraction of 
    $30$~percent of the original learning data. As it can be expected that the lagged

@@

To compare the predictive performance of all proposed models, a circular
analogue of the continuous ranked probability score (CRPS) as introduced by
\cite{Grimit+Gneiting+Berrocal:2006} is computed. Just as the linear version of
the CRPS \citep[for more details see][]{Hersbach:2000} it is a proper scoring
rule \citep{Gneiting+Raftery:2007} and measures the difference between an
observation and the corresponding predicted distribution function in order to
assess the probabilistic goodness of fit for the estimated model. Hence, the
lower the CRPS value the better the predictive performance. Contrary to the
linear version, the circular CRPS reduces not to the absolute error but to the
angular distance when the forecast is deterministic.

In addition to the raw CRPS, corresponding skill scores are computed to
assess differences in the improvement of the various statistical models over
the climatological model used as a reference:
%%
\begin{equation} 
  \text{CRPSS}_{\text{model}} = 1 -
  \frac{\text{CRPS}_{\text{model}}}{\text{CRPS}_{\text{climatology}}}.
\end{equation}
%%

All scores presented in the next section are computed out-of-sample based on
five years of data. For the persistency model only time points prior to
the time of interest are used and the validation is performed rolling over all
observations. For all other models a five-fold cross-validation is employed
using up to four calendar years for model training and the remaining single
calendar year for validation. Due to the large sample size of 24 hourly values
per day over five years, some kind of temporal aggregation is needed to ensure
a correct visual comparison of the individual methods. The analyses performed have
shown that for the employed models the variability of the predictive
performance over the five years is lower than over a single day or over a
single year. Hence, CRPS and CRPS skill scores are aggregated over the
respective five validation years which yields 24 hourly scores per month
averaged over the five validation years.

\subsection{Results} 
This section provides a detailed analysis on the predictive performance of the
different proposed statistical models applied to probabilistic wind direction
forecasting. To ensure a comprehensive comparison of the models, wind direction
forecasts are evaluated for two different lead times at two airports with
different climatological site characteristics.
Figure~\ref{fig:boxplot_crpsraw} shows the CRPS values of the employed models
at forecast steps 1\,h and 3\,h for the airports Innsbruck (Panels~a,\,c) and
Vienna (Panels~b,\,d). The scores are aggregated over the five validation years,
yielding yearly mean values for every hour per calendar month, with a lower
score indicating better performance. For both stations at both forecast steps,
the circular forest overall provides the best predictive performance, followed
by the circular regression tree and the persistency model; except for the $3$\,h~forecast 
at Innsbruck, where the persistency model is outperformed by all others. 
In comparison to the circular regression tree and
forest, for both stations and forecast steps, the climatological model and the
linear model show clearly higher CRPS values and hence a lower predictive
performance.
%%
\begin{figure}[t]
  \centering
  \setkeys{Gin}{width=1.0\textwidth}
  \includegraphics{_plot_circforest_validation_crpsraw_agg_comparison_with_lowff_v14b.pdf}
  \caption{CRPS skill scores of wind direction forecasts based on the full
    predictive von~Mises distribution for $+1$ and $+3$\,h forecasts at the
    airports Innsbruck and Vienna. Each box-and-whisker contains $24$~hourly scores
    for each of the $12$~months averaged over the $5$~validation years which
    yields a total of $288$~yearly mean values. The scores are shown for the
    climatological, persistency and the linear model as well as for the circular
    distributional tree and forest.}
  \label{fig:boxplot_crpsraw} 
\end{figure}
%%
The different site characteristics of the airports Innsbruck
(Figure~\ref{fig:boxplot_crpsraw}\,a,\,c) and Vienna
(Figure~\ref{fig:boxplot_crpsraw}\,b,\,d) seem to have an effect on the
absolute level of the model performances and on their respective predictive
performance variances. At Innsbruck, due to the surrounding mountains only a
limited number of possible wind directions exists, namely the three wind
regimes discussed for Figure~\ref{fig:tree_ibk} in
Section~\ref{sec:introduction}. Therefore, for Innsbruck the wind direction
remains rather constant in one of these possible states, but once a change
takes place it is mostly a major wind direction shift, 
e.g., from up-valley to down-valley. Due to the few wind
regimes the rather inflexible climatological and linear models score relatively
well with similar CRPS values as the other models
(Figure~\ref{fig:boxplot_crpsraw}\,a,\,c). In addition, at Innsbruck the
potential high prediction errors in case of a change of the wind regime seem to 
lead to a higher variation in the predictive performance for all models in comparison to
Vienna; this variation is especially high for the persistency model due to its
strong vulnerability to abrupt wind shifts. On the contrary, at Vienna the
geographic surroundings do not constrain any wind direction which leads to
smaller and less abrupt changes in the wind direction as well as less
pronounced wind regimes. This seems to weaken the predictive performance of the
climatological and linear models, and to reduce the performance variability for
all models (Figure~\ref{fig:boxplot_crpsraw}\,b,\,d).

The different forecast steps have apparently only a minor effect on the
predictive performance of the climatological model and the linear model at both
stations. As expected, for the persistency model, at both stations, higher scores
for the $3$\,h forecast (Figure~\ref{fig:boxplot_crpsraw}\,c,\,d) reveal a
lower performance for longer lead times; this is due to the lower 
information content of minimum 3-hourly instead of 1-hourly lagged response 
values employed as covariates in the persistency model. 
The circular regression tree and forest seem to
partially compensate for the lower skill of the lagged response values by other
covariates, hence their predictive performance only slightly decreases for the
longer lead time. This compensation is especially evident for Innsbruck, where
the performance difference between the persistency model and the tree-based
methods significantly increases from the 1-hourly to the 3-hourly forecast.

In addition to the raw CRPS (Figure~\ref{fig:boxplot_crpsraw}), CRPS skill
scores with the climatological model as a reference are provided in
Figure~\ref{fig:boxplot_crpsskill}. Skill scores are in percent, where positive
values indicate an improvement in the predictive performance over the
reference. For all setups, the circular forest has the highest skill scores
with a mean performance gain of 13--25\% and 58--71\% for Innsbruck and Vienna,
respectively. As discussed for Figure~\ref{fig:boxplot_crpsraw}, this
improvement over the climatological model is lower for Innsbruck due to the low
number of predominant wind regimes and hence a relatively good performance of
the climatological model. Additionally, Figure~\ref{fig:boxplot_crpsskill}
shows that while the persistency model's performance is lower than the
reference (Panel~c) the tree-based models can compensate for the low skill of the 
lagged response values employed as covariates and, hence, are still significantly 
superior to the reference.

@@

% -------------------------------------------------------------------
% Summary
% -------------------------------------------------------------------
\section{Summary and conclusion}
\label{sec:summary}
Circular regression trees and forests for circular responses are established by coupling 
model-based recursive partitioning with the von Mises distribution.

The resulting 
trees can capture non-linear changes, shifts, and potential interactions in covariates 
without prespecification of such effects. This is particularly useful for modeling a 
highly fluctuating response such as wind direction. \dots
